# Sparse and Ragged Data Structures  {#sparse-ragged.chapter}

Stan does not directly support either sparse or ragged data
structures, though both can be accommodated with some programming
effort.  The [sparse matrices chapter](#sparse-matrices.chapter)
introduces a special-purpose sparse matrix times dense vector
multiplication, which should be used where applicable; this chapter
covers more general data structures.

## Sparse data structures

Coding sparse data structures is as easy as moving from a matrix-like
data structure to a database-like data structure.  For example,
consider the coding of sparse data for the IRT models discussed in the
[item-response model section](#item-response-models.section).  There
are $J$ students and $K$ questions, and if every student answers every
question, then it is practical to declare the data as a $J \times K$
array of answers.

```stan
data {
  int<lower=1> J;
  int<lower=1> K;
  array[J, K] int<lower=0, upper=1> y;
  ...
model {
  for (j in 1:J) {
    for (k in 1:K) {
      y[j, k] ~ bernoulli_logit(delta[k] * (alpha[j] - beta[k]));
    }
  }
  ...
```

```{r results="asis"}
cat(ifelse(knitr::is_html_output(), "<table><tr><td>",
           "\\begin{figure}
            \\begin{center}
            \\begin{minipage}[c]{0.45\\textwidth}"))
```
$$
y
=
\left[
\begin{array}{cccc}
0 & 1 & \mbox{NA} & 1
\\
0 & \mbox{NA} & \mbox{NA} & 1
\\
\mbox{NA} & 0 & \mbox{NA} & \mbox{NA}
\end{array}
\right]
$$
```{r results="asis"}
cat(ifelse(knitr::is_html_output(), "</td><td>",
           "\\end{minipage}
            \\ \\ \\
            \\begin{minipage}[c]{0.45\\textwidth}"))
```

```{r}
df <- read.table(text=
"$jj$ | $kk$ | $y$
 1    | 1    | 0
 1    | 2    | 1
 1    | 4    | 1
 2    | 1    | 0
 2    | 4    | 1
 3    | 2    | 0
", sep="|", header=TRUE, check.names=FALSE)
kable(df, booktabs=TRUE, escape=FALSE, linesep="") %>%
  kable_styling(full_width=TRUE)
```

```{r results="asis"}
cat(ifelse(knitr::is_html_output(), "</td></tr></table>",
           "\\end{minipage}
            \\end{center}
            \\end{figure}"))
```

On the left is a definition of a sparse matrix $y$ using the NA notation from R
(which is not supported by Stan).  On the right is a database-like encoding of
the same sparse matrix $y$ that can be used directly in Stan.  The first two
columns, $jj$ and $kk$, denote the indexes and the final column, $y$, the value.
For example, the fifth row of the database-like data structure on the right
indicates that $y_{2,4} = 1$.


When not every student is given every question, the dense array coding
will no longer work, because Stan does not support undefined values.
The sparse data example shows an example with $J=3$ and $K=4$, with
missing responses shown as NA, as in R.  There is no support within
Stan for R's NA values, so this data structure cannot be used
directly.  Instead, it must be converted to a "long form" as in a
database, with columns indicating the $j$ and $k$ indexes along with
the value.  For instance, with $jj$ and $kk$ used for the indexes
(following @GelmanHill:2007), the data structure can be coded as in
the right-hand example in the example. This says that
$y_{1,1} = 0$, $y_{1,2} = 1$, and so on, up to $y_{3,2} = 1$, with all
other entries undefined.

Letting $N$ be the number of $y$ that are defined, here $N=6$,
the data and model can be formulated as follows.

```
data {
  ...
  int<lower=1> N;
  array[N] int<lower=1, upper=J> jj;
  array[N] int<lower=1, upper=K> kk;
  array[N] int<lower=0, upper=1> y;
  ...
model {
  for (n in 1:N) {
    y[n] ~ bernoulli_logit(delta[kk[n]]
                           * (alpha[jj[n]] - beta[kk[n]]));
  }
  ...
```

In the situation where there are no missing values, the two model
formulations produce exactly the same log posterior density.


## Ragged data structures {#ragged-data-structs.section}

Ragged arrays are arrays that are not rectangular, but have different
sized entries.  This kind of structure crops up when there are
different numbers of observations per entry.

A general approach to dealing with ragged structure is to move to a
full database-like data structure as discussed in the previous
section.  A more compact approach is possible with some indexing into
a linear array.

For example, consider a data structure for three groups, each of which
has a different number of observations.

```{r results="asis"}
cat(ifelse(knitr::is_html_output(), "<table><tr><td>",
           "\\begin{center}
            \\begin{minipage}[c]{0.35\\textwidth}"))
```
$y_1 =  \left[1.3 \ \ 2.4 \ \ 0.9\right]$

$y_2 = \left[-1.8 \ \ -0.1\right]$

$y_3 = \left[12.9 \ \ 18.7 \ \ 42.9 \ \ 4.7\right]$

```{r results="asis"}
cat(ifelse(knitr::is_html_output(), "</td><td>",
           "\\end{minipage}
            \\ \\ \\
            \\begin{minipage}[c]{0.60\\textwidth}"))
```

$z = [1.3 \ \ 2.4 \ \ 0.9 \ \ -1.8 \ \ -0.1 \ \ 12.9 \ \ 18.7 \ \ 42.9
\ \ 4.7]$

$s  =  \{ 3 \ \ 2 \ \ 4 \}$


```{r results="asis"}
cat(ifelse(knitr::is_html_output(), "</td></tr></table>",
           "\\end{minipage}
            \\end{center}"))
```

On the left is the definition of a ragged data structure $y$ with three rows of
different sizes ($y_1$ is size 3, $y_2$ size 2, and $y_3$ size 4).  On the right
is an example of how to code the data in Stan, using a single vector $y$ to hold
all the values and a separate array of integers $s$ to hold the group row sizes.
In this example, $y_1 = z_{1:3}$, $y_2 = z_{4:5}$, and $y_3 = z_{6:9}$.

Suppose the model is a  simple varying intercept model, which,
using vectorized notation, would yield a log-likelihood
$$
\sum_{n=1}^3 \log \textsf{normal}(y_n \mid \mu_n, \sigma).
$$
There's no direct way to encode this in Stan.

A full database type structure could be used, as in the sparse
example, but this is inefficient, wasting space for unnecessary
indices and not allowing vector-based density operations.  A better
way to code this data is as a single list of values, with a separate
data structure indicating the sizes of each subarray.  This is
indicated on the right of the example. This coding uses a
single array for the values and a separate array for the sizes of each
row.

The model can then be coded up using slicing operations as follows.

```
data {
  int<lower=0> N;   // # observations
  int<lower=0> K;   // # of groups
  vector[N] y;      // observations
  array[K] int s;   // group sizes
  ...
model {
  int pos;
  pos = 1;
  for (k in 1:K) {
    segment(y, pos, s[k]) ~ normal(mu[k], sigma);
    pos = pos + s[k];
  }
```

This coding allows for efficient vectorization, which is worth the
copy cost entailed by the `segment()` vector slicing operation.
