# Bounded Continuous Distributions
```{r include, file="_functions.R", echo=FALSE}
```


The bounded continuous probabilities have support on a finite interval
of real numbers.

```{r results='asis', echo=FALSE}
if (knitr::is_html_output()) {
cat(' * <a href="uniform-distribution.html">Uniform Distribution</a>\n')
}
```

## Uniform distribution

### Probability density function

If $\alpha \in \mathbb{R}$ and $\beta \in (\alpha,\infty)$, then for
$y \in [\alpha,\beta]$, \begin{equation*} \text{Uniform}(y|\alpha,\beta) =
\frac{1}{\beta - \alpha} . \end{equation*}

### Sampling statement

`y ~ ` **`uniform`**`(alpha, beta)`

Increment target log probability density with `uniform_lupdf(y | alpha, beta)`.
`r since("2.0")`
<!-- real; uniform ~; -->
\index{{\tt \bfseries uniform }!sampling statement|hyperpage}

### Stan functions

<!-- real; uniform_lpdf; (reals y | reals alpha, reals beta); -->
\index{{\tt \bfseries uniform\_lpdf }!{\tt (reals y \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`uniform_lpdf`**`(reals y | reals alpha, reals beta)`<br>\newline
The log of the uniform density of y given lower bound alpha and upper
bound beta
`r since("2.12")`

<!-- real; uniform_lupdf; (reals y | reals alpha, reals beta); -->
\index{{\tt \bfseries uniform\_lupdf }!{\tt (reals y \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`uniform_lupdf`**`(reals y | reals alpha, reals beta)`<br>\newline
The log of the uniform density of y given lower bound alpha and upper
bound beta dropping constant additive terms
`r since("2.25")`

<!-- real; uniform_cdf; (reals y, reals alpha, reals beta); -->
\index{{\tt \bfseries uniform\_cdf }!{\tt (reals y, reals alpha, reals beta): real}|hyperpage}

`real` **`uniform_cdf`**`(reals y, reals alpha, reals beta)`<br>\newline
The uniform cumulative distribution function of y given lower bound
alpha and upper bound beta
`r since("2.0")`

<!-- real; uniform_lcdf; (reals y | reals alpha, reals beta); -->
\index{{\tt \bfseries uniform\_lcdf }!{\tt (reals y \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`uniform_lcdf`**`(reals y | reals alpha, reals beta)`<br>\newline
The log of the uniform cumulative distribution function of y given
lower bound alpha and upper bound beta
`r since("2.12")`

<!-- real; uniform_lccdf; (reals y | reals alpha, reals beta); -->
\index{{\tt \bfseries uniform\_lccdf }!{\tt (reals y \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`uniform_lccdf`**`(reals y | reals alpha, reals beta)`<br>\newline
The log of the uniform complementary cumulative distribution function
of y given lower bound alpha and upper bound beta
`r since("2.12")`

<!-- R; uniform_rng; (reals alpha, reals beta); -->
\index{{\tt \bfseries uniform\_rng  }!{\tt (reals alpha, reals beta): R}|hyperpage}

`R` **`uniform_rng`**`(reals alpha, reals beta)`<br>\newline
Generate a uniform variate with lower bound alpha and upper bound
beta; may only be used in transformed data and generated quantities blocks. For a
description of argument and return types, see section
[vectorized PRNG functions](#prng-vectorization).
`r since("2.18")`
