# Continuous Distributions on [0, 1]
```{r include, file="_functions.R", echo=FALSE}

```

The continuous distributions with outcomes in the interval $[0,1]$ are
used to characterized bounded quantities, including probabilities.

```{r results='asis', echo=FALSE}
if (knitr::is_html_output()) {
cat(' * <a href="beta-distribution.html">Beta Distribution</a>\n')
cat(' * <a href="beta-proportion-distribution.html">Beta Proportion Distribution</a>\n')
}
```

## Beta distribution

### Probability density function

If $\alpha \in \mathbb{R}^+$ and $\beta \in \mathbb{R}^+$, then for
$\theta \in (0,1)$, \begin{equation*} \text{Beta}(\theta|\alpha,\beta) =
\frac{1}{\mathrm{B}(\alpha,\beta)} \, \theta^{\alpha - 1} \, (1 -
\theta)^{\beta - 1} , \end{equation*} where the beta function $\mathrm{B}()$ is as
defined in section [combinatorial functions](#betafun).

_**Warning:**_  If $\theta = 0$ or $\theta = 1$, then the probability
is 0 and the log probability is $-\infty$.  Similarly, the
distribution requires strictly positive parameters, $\alpha, \beta >
0$.

### Sampling statement

`theta ~ ` **`beta`**`(alpha, beta)`

Increment target log probability density with `beta_lupdf(theta | alpha, beta)`.
`r since("2.0")`
<!-- real; beta ~; -->
\index{{\tt \bfseries beta }!sampling statement|hyperpage}

### Stan functions

<!-- real; beta_lpdf; (reals theta | reals alpha, reals beta); -->
\index{{\tt \bfseries beta\_lpdf }!{\tt (reals theta \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`beta_lpdf`**`(reals theta | reals alpha, reals beta)`<br>\newline
The log of the beta density of `theta` in $[0,1]$ given positive prior
successes (plus one) alpha and prior failures (plus one) beta
`r since("2.12")`

<!-- real; beta_lupdf; (reals theta | reals alpha, reals beta); -->
\index{{\tt \bfseries beta\_lupdf }!{\tt (reals theta \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`beta_lupdf`**`(reals theta | reals alpha, reals beta)`<br>\newline
The log of the beta density of `theta` in $[0,1]$ given positive prior
successes (plus one) alpha and prior failures (plus one) beta
dropping constant additive terms
`r since("2.25")`

<!-- real; beta_cdf; (reals theta | reals alpha, reals beta); -->
\index{{\tt \bfseries beta\_cdf }!{\tt (reals theta  \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`beta_cdf`**`(reals theta | reals alpha, reals beta)`<br>\newline
The beta cumulative distribution function of `theta` in $[0,1]$ given
positive prior successes (plus one) alpha and prior failures (plus
one) beta
`r since("2.0")`

<!-- real; beta_lcdf; (reals theta | reals alpha, reals beta); -->
\index{{\tt \bfseries beta\_lcdf }!{\tt (reals theta \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`beta_lcdf`**`(reals theta | reals alpha, reals beta)`<br>\newline
The log of the beta cumulative distribution function of `theta` in
$[0,1]$ given positive prior successes (plus one) alpha and prior
failures (plus one) beta
`r since("2.12")`

<!-- real; beta_lccdf; (reals theta | reals alpha, reals beta); -->
\index{{\tt \bfseries beta\_lccdf }!{\tt (reals theta \textbar\ reals alpha, reals beta): real}|hyperpage}

`real` **`beta_lccdf`**`(reals theta | reals alpha, reals beta)`<br>\newline
The log of the beta complementary cumulative distribution function of
`theta` in $[0,1]$ given positive prior successes (plus one) alpha and
prior failures (plus one) beta
`r since("2.12")`

<!-- R; beta_rng; (reals alpha, reals beta); -->
\index{{\tt \bfseries beta\_rng }!{\tt (reals alpha, reals beta): R}|hyperpage}

`R` **`beta_rng`**`(reals alpha, reals beta)`<br>\newline
Generate a beta variate with positive prior successes (plus one) alpha
and prior failures (plus one) beta; may only be used in transformed data and
generated quantities blocks. For a description of argument and return types, see
section [vectorized PRNG functions](#prng-vectorization).
`r since("2.18")`

## Beta proportion distribution

### Probability density function

If $\mu \in (0, 1)$ and $\kappa \in \mathbb{R}^+$, then for $\theta
\in (0,1)$, \begin{equation*} \mathrm{Beta\_Proportion}(\theta|\mu,\kappa) =
\frac{1}{\mathrm{B}(\mu \kappa, (1 - \mu) \kappa)} \,
\theta^{\mu\kappa - 1} \, (1 - \theta)^{(1 - \mu)\kappa- 1} , \end{equation*} where
the beta function $\mathrm{B}()$ is as defined in section
[combinatorial functions](#betafun).

_**Warning:**_  If $\theta = 0$ or $\theta = 1$, then the probability
is 0 and the log probability is $-\infty$.  Similarly, the
distribution requires $\mu \in (0, 1)$ and strictly positive
parameter, $\kappa > 0$.

### Sampling statement

`theta ~ ` **`beta_proportion`**`(mu, kappa)`

Increment target log probability density with `beta_proportion_lupdf(theta | mu, kappa)`.
`r since("2.19")`
<!-- real; beta_proportion ~; -->
\index{{\tt \bfseries beta\_proportion }!sampling statement|hyperpage}

### Stan functions

<!-- real; beta_proportion_lpdf; -->
\index{{\tt \bfseries beta\_proportion\_lpdf }!{\tt (reals theta \textbar\ reals mu, reals kappa): real}|hyperpage}

`real` **`beta_proportion_lpdf`**`(reals theta | reals mu, reals kappa)`<br>\newline
The log of the beta_proportion density of `theta` in $(0,1)$ given
mean mu and precision kappa
`r since("2.19")`

<!-- real; beta_proportion_lupdf; -->
\index{{\tt \bfseries beta\_proportion\_lupdf }!{\tt (reals theta \textbar\ reals mu, reals kappa): real}|hyperpage}

`real` **`beta_proportion_lupdf`**`(reals theta | reals mu, reals kappa)`<br>\newline
The log of the beta_proportion density of `theta` in $(0,1)$ given
mean mu and precision kappa dropping constant additive terms
`r since("2.25")`

<!-- real; beta_proportion_lcdf; (reals theta | reals mu, reals kappa); -->
\index{{\tt \bfseries beta\_proportion\_lcdf }!{\tt (reals theta \textbar\ reals mu, reals kappa): real}|hyperpage}

`real` **`beta_proportion_lcdf`**`(reals theta | reals mu, reals kappa)`<br>\newline
The log of the beta_proportion cumulative distribution function of
`theta` in $(0,1)$ given mean mu and precision kappa
`r since("2.18")`

<!-- real; beta_proportion_lccdf; (reals theta | reals mu, reals kappa); -->
\index{{\tt \bfseries beta\_proportion\_lccdf }!{\tt (reals theta \textbar\ reals mu, reals kappa): real}|hyperpage}

`real` **`beta_proportion_lccdf`**`(reals theta | reals mu, reals kappa)`<br>\newline
The log of the beta_proportion complementary cumulative distribution
function of `theta` in $(0,1)$ given mean mu and precision kappa
`r since("2.18")`

<!-- R; beta_proportion_rng; (reals mu, reals kappa); -->
\index{{\tt \bfseries beta\_proportion\_rng }!{\tt (reals mu, reals kappa): R}|hyperpage}

`R` **`beta_proportion_rng`**`(reals mu, reals kappa)`<br>\newline
Generate a beta_proportion variate with mean mu and precision kappa;
may only be used in transformed data and generated quantities blocks.
For a description of argument and return types, see section
[vectorized PRNG functions](#prng-vectorization).
`r since("2.18")`
