# Integer-Valued Basic Functions
```{r include, file="_functions.R", echo=FALSE}
```


This chapter describes Stan's built-in function that take various
types of arguments and return results of type integer.

```{r results='asis', echo=FALSE}
if (knitr::is_html_output()) {
cat(' * <a href="int-arithmetic.html">Integer-Valued Arithmetic Operators</a>\n')
cat(' * <a href="absolute-functions.html">Absolute Functions</a>\n')
cat(' * <a href="bound-functions.html">Bound Functions</a>\n')
cat(' * <a href="size-functions.html">Size Functions</a>\n')
cat(' * <a href="casting-functions.html">Casting Functions</a>\n')
}
```

## Integer-valued arithmetic operators {#int-arithmetic}

Stan's arithmetic is based on standard double-precision  C++ integer
and floating-point arithmetic.  If the arguments to an arithmetic
operator are both integers, as in `2 + 2`, integer arithmetic is used.
If one argument is an integer and the other a floating-point value, as
in `2.0 + 2` and `2 + 2.0`, then the integer is promoted to a floating
point value and floating-point arithmetic is used.

Integer arithmetic behaves slightly differently than floating point
arithmetic.  The first difference is how overflow is treated.  If the
sum or product of two integers overflows the maximum integer
representable, the result is an undesirable wraparound behavior at the
bit level.  If the integers were first promoted to real numbers, they
would not overflow a floating-point representation.  There are no
extra checks in Stan to flag overflows, so it is up to the user to
make sure it does not occur.

Secondly, because the set of integers is not closed under division and
there is no special infinite value for integers, integer division
implicitly rounds the result.  If both arguments are positive, the
result is rounded down.  For example, `1 / 2` evaluates to 0 and `5 /
3` evaluates to 1.

If one of the integer arguments to division is negative, the latest
C++ specification ( C++11), requires rounding toward zero.  This would
have `1 / 2` and `-1 / 2` evaluate to 0, `-7 / 2` evaluate to -3, and
`7 / 2` evaluate to 3.  Before the  C++11 specification, the behavior
was platform dependent, allowing rounding up or down.  All compilers
recent enough to be able to deal with Stan's templating should follow
the  C++11 specification, but it may be worth testing if you are not
sure and plan to use integer division with negative values.

Unlike floating point division, where `1.0 / 0.0` produces the special
positive infinite value, integer division by zero, as in `1 / 0`, has
undefined behavior in the  C++ standard.  For example, the clang++
compiler on Mac OS X returns 3764, whereas the g++ compiler throws an
exception and aborts the program with a warning.  As with overflow, it
is up to the user to make sure integer divide-by-zero does not occur.

### Binary infix operators

Operators are described using the  C++ syntax.  For instance, the
binary operator of addition, written `X + Y`, would have the Stan
signature `int operator+(int, int)` indicating it takes two real
arguments and returns a real value.  As noted previously, the value of
integer division is platform-dependent when rounding is platform
dependent before C++11;  the descriptions below provide the C++11
definition.

<!-- int; operator+; (int x, int y); -->
\index{{\tt \bfseries operator\_add }!{\tt (int x, int y): int}|hyperpage}

`int` **`operator+`**`(int x, int y)`<br>\newline
The sum of the addends x and y \begin{equation*} \text{operator+}(x,y) = (x + y) \end{equation*}
`r since("2.0")`

<!-- int; operator-; (int x, int y); -->
\index{{\tt \bfseries operator\_subtract }!{\tt (int x, int y): int}|hyperpage}

`int` **`operator-`**`(int x, int y)`<br>\newline
The difference between the minuend x and subtrahend y \begin{equation*}
\text{operator-}(x,y) = (x - y) \end{equation*}
`r since("2.0")`

<!-- int; operator*; (int x, int y); -->
\index{{\tt \bfseries operator\_multiply }!{\tt (int x, int y): int}|hyperpage}

`int` **`operator*`**`(int x, int y)`<br>\newline
The product of the factors x and y \begin{equation*} \text{operator*}(x,y) = (x
\times y) \end{equation*}
`r since("2.0")`

<!-- int; operator/; (int x, int y); -->
\index{{\tt \bfseries operator\_divide }!{\tt (int x, int y): int}|hyperpage}

`int` **`operator/`**`(int x, int y)`<br>\newline
The integer quotient of the dividend x and divisor y \begin{equation*}
\text{operator/}(x,y) = \begin{cases} \lfloor x / y \rfloor & \text{if
} x / y \geq 0 \\ - \lfloor \text{floor}(-x / y) \rfloor & \text{if }
x / y < 0. \end{cases} \end{equation*}
**deprecated;** - use `operator%/%` instead.
`r since("2.0, deprecated in 2.24")`


<!-- int; operator%/%; (int x, int y); -->
\index{{\tt \bfseries operator\_int\_divide }!{\tt (int x, int y): int}|hyperpage}

`int` **`operator%/%`**`(int x, int y)`<br>\newline
The integer quotient of the dividend x and divisor y \begin{equation*}
\text{operator\%/\%}(x,y) = \begin{cases} \lfloor x / y \rfloor & \text{if
} x / y \geq 0 \\ - \lfloor \text{floor}(-x / y) \rfloor & \text{if }
x / y < 0. \end{cases} \end{equation*}
`r since("2.24")`

<!-- int; operator%; (int x, int y); -->
\index{{\tt \bfseries operator\_mod }!{\tt (int x, int y): int}|hyperpage}

`int` **`operator%`**`(int x, int y)`<br>\newline
x modulo y, which is the positive remainder after dividing x by y. If
both x and y are non-negative, so is the result; otherwise, the sign
of the result is platform dependent. \begin{equation*} \mathrm{operator\%}(x, y) \ =
\ x \ \text{mod} \ y \ = \ x - y * \lfloor x / y \rfloor \end{equation*}
`r since("2.13")`

### Unary prefix operators

<!-- int; operator-; (int x); -->
\index{{\tt \bfseries operator\_subtract }!{\tt (int x): int}|hyperpage}

`int` **`operator-`**`(int x)`<br>\newline
The negation of the subtrahend x \begin{equation*} \text{operator-}(x) = -x \end{equation*}
`r since("2.0")`

<!-- T; operator-; (T x); -->
\index{{\tt \bfseries operator\_subtract }!{\tt (T x): T}|hyperpage}

`T` **`operator-`**`(T x)`<br>\newline
Vectorized version of `operator-`. If `T x` is a (possibly nested) array of
integers, `-x` is the same shape array where each individual integer is negated.
`r since("2.31")`


<!-- int; operator+; (int x); -->
\index{{\tt \bfseries operator\_add }!{\tt (int x): int}|hyperpage}

`int` **`operator+`**`(int x)`<br>\newline
This is a no-op. \begin{equation*} \text{operator+}(x) = x \end{equation*}
`r since("2.0")`

## Absolute functions

<!-- T; abs; (T x); -->
\index{{\tt \bfseries abs }!{\tt (T x): T}|hyperpage}

`T` **`abs`**`(T x)`<br>\newline
The absolute value of x.

This function works elementwise over containers such as vectors.
Given a type `T` which is `int`, or an array of `int`s, `abs` returns the same
type where each element has had its absolute value taken.
`r since("2.0, vectorized in 2.30")`

<!-- int; int_step; (int x); -->
\index{{\tt \bfseries int\_step }!{\tt (int x): int}|hyperpage}

`int` **`int_step`**`(int x)`<br>\newline

<!-- int; int_step; (real x); -->
\index{{\tt \bfseries int\_step }!{\tt (real x): int}|hyperpage}

`int` **`int_step`**`(real x)`<br>\newline
Return the step function of x as an integer, \begin{equation*} \mathrm{int\_step}(x)
= \begin{cases} 1 & \text{if } x > 0 \\ 0 & \text{if } x \leq 0 \text{
or } x \text{ is } NaN \end{cases} \end{equation*} _**Warning:**_ `int_step(0)` and
`int_step(NaN)` return 0 whereas `step(0)` and `step(NaN)` return 1.

See the warning in section [step functions](#step-functions) about the dangers of
step functions applied to anything other than data.
`r since("2.0")`

## Bound functions

<!-- int; min; (int x, int y); -->
\index{{\tt \bfseries min }!{\tt (int x, int y): int}|hyperpage}

`int` **`min`**`(int x, int y)`<br>\newline
Return the minimum of x and y. \begin{equation*} \text{min}(x, y) = \begin{cases} x &
\text{if } x < y\\ y & \text{otherwise} \end{cases} \end{equation*}
`r since("2.0")`

<!-- int; max; (int x, int y); -->
\index{{\tt \bfseries max }!{\tt (int x, int y): int}|hyperpage}

`int` **`max`**`(int x, int y)`<br>\newline
Return the maximum of x and y. \begin{equation*} \text{max}(x, y) = \begin{cases} x &
\text{if } x > y\\ y & \text{otherwise} \end{cases} \end{equation*}
`r since("2.0")`

## Size functions

<!-- int; size; (int x); -->
\index{{\tt \bfseries size }!{\tt (int x): int}|hyperpage}

`int` **`size`**`(int x)`<br>\newline

<!-- int; size; (real x); -->
\index{{\tt \bfseries size }!{\tt (real x): int}|hyperpage}

`int` **`size`**`(real x)`<br>\newline

Return the size of `x` which for scalar-valued `x` is 1
`r since("2.26")`

## Casting functions

It is possible to cast real numbers to integers as long as the real
value is `data`. See
[data only qualifiers](https://mc-stan.org/docs/reference-manual/argument-types-and-qualifiers.html#data-only-qualifiers)
in the Stan Reference Manual.

<!-- int; to_int; (data real x); -->
\index{{\tt \bfseries to\_int }!{\tt (data real x): int}|hyperpage}

`int` **`to_int`**`(data real x)`<br>\newline

Return the value `x` truncated to an integer. This will throw an error
if the value of `x` is too big to represent as a 32-bit signed integer.

This is similar to `trunc` (see [Rounding functions]) but the return type is of
type `int`. For example, `to_int(3.9)` is `3`, and `to_int(-3.9)` is `-3`.
`r since("2.31")`


<!-- int; to_int; (data real x); -->
\index{{\tt \bfseries to\_int }!{\tt (data R x): I}|hyperpage}

`I` **`to_int`**`(data T x)`<br>\newline

The vectorized version of `to_int`. This function accepts a (possibly nested)
array of reals and returns an array of the same shape where each element has
been truncated to an integer.
`r since("2.31")`
