# Sparse Matrix Operations {#sparse-matrices}
```{r include, file="_functions.R", echo=FALSE}
```


For sparse matrices, for which many elements are zero, it is more
efficient to use specialized representations to save memory and speed
up matrix arithmetic (including derivative calculations).  Given
Stan's implementation, there is substantial space (memory) savings by
using sparse matrices.  Because of the ease of optimizing dense matrix
operations, speed improvements only arise at 90\% or even greater
sparsity; below that level, dense matrices are faster but use more
memory.

Because of this speedup and space savings, it may even be useful to
read in a dense matrix and convert it to a sparse matrix before
multiplying it by a vector.  This chapter covers a very specific form
of sparsity consisting of a sparse matrix multiplied by a dense
vector.

```{r results='asis', echo=FALSE}
if (knitr::is_html_output()) {
cat(' * <a href="CSR.html">Compressed Row Storage</a>\n')
cat(' * <a href="conversion-functions.html">Conversion Functions</a>\n')
cat(' * <a href="sparse-matrix-arithmetic.html">Sparse Matrix Arithmetic</a>\n')
}
```

## Compressed row storage {#CSR}

Sparse matrices are represented in Stan using compressed row storage
(CSR).  For example, the matrix
\begin{equation*}
A = \begin{bmatrix} 19 & 27 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 52 \\ 81 & 0 & 95 & 33 \end{bmatrix}
\end{equation*}
is translated into a vector of the non-zero real values, read by row from the matrix $A$,
\begin{equation*}
w(A) = \begin{bmatrix} 19 & 27 & 52 & 81 & 95 & 33 \end{bmatrix}^{\top} \! \! \! ,
\end{equation*}
an array of integer column indices for the values,
\begin{equation*}
v(A) = \begin{bmatrix} 1 & 2 & 4 & 1 & 3 & 4 \end{bmatrix} \! ,
\end{equation*}
and an array of integer indices indicating where in $w(A)$ a given row's values start,
\begin{equation*} u(A) = \begin{bmatrix} 1 & 3 & 3 & 4 & 7 \end{bmatrix} \! ,
\end{equation*}
with a padded value at the end to guarantee that
\begin{equation*} u(A)[n+1] - u(A)[n] \end{equation*}
is the number of non-zero elements in row $n$ of the matrix (here $2$, $0$, $1$, and $3$).
Note that because the second row has no non-zero elements both the second
and third elements of $u(A)$ correspond to the third element of
$w(A)$, which is $52$. The values $(w(A), \, v(A), \, u(A))$ are
sufficient to reconstruct $A$.

The values are structured so that there is a real value and integer
column index for each non-zero entry in the array, plus one integer
for each row of the matrix, plus one for padding.  There is also
underlying storage for internal container pointers and sizes.  The
total memory usage is roughly $12 K + M$ bytes plus a small constant
overhead, which is often considerably fewer bytes than the $M \times
N$ required to store a dense matrix.  Even more importantly, zero
values do not introduce derivatives under multiplication or addition,
so many storage and evaluation steps are saved when sparse matrices
are multiplied.

## Conversion functions

Conversion functions between dense and sparse matrices are provided.

### Dense to sparse conversion

Converting a dense matrix $m$ to a sparse representation produces a
vector $w$ and two integer arrays, $u$ and $v$.

<!-- vector; csr_extract_w; (matrix a); -->
\index{{\tt \bfseries csr\_extract\_w }!{\tt (matrix a): vector}|hyperpage}

`vector` **`csr_extract_w`**`(matrix a)`<br>\newline
Return non-zero values in matrix a; see section [compressed row storage](#CSR).
`r since("2.8")`

<!-- array[] int; csr_extract_v; (matrix a); -->
\index{{\tt \bfseries csr\_extract\_v }!{\tt (matrix a): array[] int}|hyperpage}

`array[] int` **`csr_extract_v`**`(matrix a)`<br>\newline
Return column indices for values in `csr_extract_w(a)`; see
[compressed row storage](#CSR).
`r since("2.8")`

<!-- array[] int; csr_extract_u; (matrix a); -->
\index{{\tt \bfseries csr\_extract\_u }!{\tt (matrix a): array[] int}|hyperpage}

`array[] int` **`csr_extract_u`**`(matrix a)`<br>\newline
Return array of row starting indices for entries in `csr_extract_w(a)`
followed by the size of `csr_extract_w(a)` plus one; see section
[compressed row storage](#CSR).
`r since("2.8")`


<!-- tuple(vector, array[] int, array[] int); csr_extract; (matrix a); -->
\index{{\tt \bfseries csr\_extract }!{\tt (matrix a): tuple(vector, array[] int, array[] int)}|hyperpage}

`tuple(vector, array[] int, array[] int)` **`csr_extract`**`(matrix a)`<br>\newline
Return all three components of the CSR representation of the matrix `a`; see
section [compressed row storage](#CSR).
This function is equivalent to `(csr_extract_w(a), csr_extract_v(a), csr_extract_u(a))`.
`r since("2.33")`

### Sparse to dense conversion

To convert a sparse matrix representation to a dense matrix, there is
a single function.

<!-- matrix; csr_to_dense_matrix; (int m, int n, vector w, array[] int v, array[] int u); -->
\index{{\tt \bfseries csr\_to\_dense\_matrix }!{\tt (int m, int n, vector w, array[] int v, array[] int u): matrix}|hyperpage}

`matrix` **`csr_to_dense_matrix`**`(int m, int n, vector w, array[] int v, array[] int u)`<br>\newline
Return dense $\text{m} \times \text{n}$ matrix with non-zero matrix
entries w, column indices v, and row starting indices u; the vector w
and array v must be the same size (corresponding to the total number of
nonzero entries in the matrix), array v must have index values bounded
by m, array u must have length equal to m + 1 and contain index values
bounded by the number of nonzeros (except for the last entry, which must
be equal to the number of nonzeros plus one). See section
[compressed row storage](#CSR) for more details.
`r since("2.10")`

## Sparse matrix arithmetic

### Sparse matrix multiplication

The only supported operation is the multiplication of a sparse matrix
$A$ and a dense vector $b$ to produce a dense vector $A\,b$.
Multiplying a dense row vector $b$ and a sparse matrix $A$ can be
coded using transposition as
\begin{equation*}
b \, A = (A^{\top} \, b^{\top})^{\top},
\end{equation*}
but care must be taken to represent $A^{\top}$ rather than $A$ as a sparse matrix.

<!-- vector; csr_matrix_times_vector; (int m, int n, vector w, array[] int v, array[] int u, vector b); -->
\index{{\tt \bfseries csr\_matrix\_times\_vector }!{\tt (int m, int n, vector w, array[] int v, array[] int u, vector b): vector}|hyperpage}

`vector` **`csr_matrix_times_vector`**`(int m, int n, vector w, array[] int v, array[] int u, vector b)`<br>\newline
Multiply the $\text{m} \times \text{n}$ matrix represented by values
w, column indices v, and row start indices u by the vector b; see
[compressed row storage](#CSR).
`r since("2.18")`
